---
title: "analysis(bmi7)"
author: "Vikram Jambunathan"
date: "4/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=FALSE, message=FALSE, warning=FALSE)
library(CpGassoc)
library(data.table)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
```

The motivation of this analysis is to analyze ALSPAC data, to run an EWAS at 9 timepoints to find out if there is any association between DNA methylation and body mass index (BMI) change during childhood and adolescence.

This project will aim to combine longitudinal data analysis approaches with standard EWAS methods to find whether including repeated measures of BMI reveals more epigenetic signatures

First step is modelling BMI over age in a group of up to 14,000 children from the Avon Longitudinal Study of Parents and Children (ALSPAC)and then perform a standard EWAS of BMI at each timepoint.

We will be using the package called CpGassoc to run the EWAS analysis.

On comparing the data that we had, the structure of betas_cord resembles samplecpg (from the example data in the CpGassoc package), and the structure of bmi_data resembles samplepheno. But there is some tidying up to do. 

We need to make our betas_cord object look EXACTLY like the samplecpg and make our bmi_data object look EXACTLY like samplepheno, link the covariates and Match the columns of betas_cord to the rows of bmi_data.

```{r}
betas_cord <- fread("betas_cord.csv")
cpgnames <- read.csv("cpgnames.csv")
bmi_data <- read.csv("bmi_data.csv")
names(bmi_data)[1] <- 'cidB3967'
link_cord <- data.frame(fread("link_cord.csv"))
link_cord <- link_cord[order(link_cord$cidB3967),]
bmi_data <- merge(link_cord,bmi_data,by="cidB3967", all.x = TRUE, all.y = TRUE)

# linking covariates to the bmi_data.

covariates <- data.frame(fread("covariates.csv"))
covariates <- covariates[order(covariates$cidB3967),]
bmi_data <- merge(bmi_data,covariates,by="cidB3967", all.x = TRUE, all.y = TRUE)

# Considering bmi7 in this file.

bmi7 <- bmi_data %>% filter(time == "F7") %>% select(Sample_Name,bmi,c(8:29)) %>% na.omit()

```
Then we will have to:

1. Remove columns of betas_cord for Sample Names not in bmi7
2. Sort bmi7 rows to be in order of the betas_cord columns
3. Change betas_cord to a matrix
4. Change row names of betas_cord matrix to cpgnames
5. Run cpg.assoc(betas_cord, bmi7$bmi)
6. Table of top 10 cpg
7. Manhattan plot

```{r}
# For bmi7, we need to remove columns of betas_cord for Sample Names not in bmi7
a <- t(bmi7)
colnames(a) <- a[1,]
a <-  as.data.frame(a[-1, ])
# a <- as.data.frame(t(a))
betas_cord <- betas_cord%>%select(which(colnames(betas_cord) %in% colnames(a)))
betas_cord <- betas_cord %>% select(sort(names(.)))

# Sorting the data so that they are in the same order

betas_cord <- betas_cord %>% select(sort(names(.)))
bmi7 <- bmi7[order(bmi7$Sample_Name),]

# Change betas_cord to a matrix
betas_cord <- as.matrix(betas_cord)
rownames(betas_cord) <- cpgnames$cpgname

# Running cpg.assoc
covar <- bmi7[c(3:24)]
cpg_assocF7 <- cpg.assoc(betas_cord, bmi7$bmi, covariates = covar)
cpg_assocF7

```
# QQ PLOT
```{r}

# Plotting the results (QQ Plot)
plot(cpg_assocF7)

```

# MANHATTAN PLOT

```{r}
# Merge with location info from the Illumina Human Methlyation Annotation Package.

cpg_assoc_results = cpg_assocF7$results
cpg_locations = data.frame(Locations)
cpg_locations$CPG.Labels = row.names(cpg_locations)
resultF7 <- merge(cpg_assoc_results, cpg_locations)
```

```{r}
cpg_anno_rn <- rownames(cpg_locations)
rownames(cpg_locations) <- NULL
cpg_locations <- cbind(cpg_anno_rn,cpg_locations)
names(cpg_locations)[1] <- 'TargetID'
cpg_locations <- cpg_locations[order(cpg_locations$TargetID),]
names(cpg_locations)[2] <- 'CHR'
names(cpg_locations)[3] <- 'MAPINFO'
cpg_locations <- cpg_locations[-c(4,5)]
cpg_locations$CHR <- gsub("chr","",as.character(cpg_locations$CHR))
rownames(cpg_locations) <- NULL
```

```{r}
# Manhattan Plot

manhattan(cpg_assocF7,cpg_locations$TargetID,cpg_locations$CHR, cpg_locations$MAPINFO)


```

#Summary Statistcs at specific timepoints.

```{r}
library(gtsummary)
library(htmlwidgets)
bmi_data_linked %>% 
  select(bmi, age, time) %>%
  tbl_summary(by = time)
reorder

```

```{r}
bmi_data_linked %>%
  filter(time == "F7" & mumsmoke!= -1) %>%
  mutate(bmi_category = cut(bmi,breaks = c(0,18.5,25,30,40))) %>%
  select(sex,bwt,gestage,mumsmoke,mumbmi,cd8t_cord,cd4t_cord,nk_cord,
              bcell_cord, mono_cord, gran_cord, bmi_category) %>%
  tbl_summary(by = bmi_category) %>%
  add_p()
```

#Boxplot of BMI vs Time (Age7~Age24) 

```{r}
bmi_data %>%
  filter(age > 6) %>%
  ggplot(aes(x = reorder(time,bmi), y = bmi, fill = time)) + labs(title = "Body Mass Index vs Time (Age7~Age 24)",x = "Time", y= "BMI") + geom_boxplot()
```








NOTE: THIS PART OF THE MARKDOWN FILE WAS ONLY USED FOR TESTING THE LONGITUDINAL ANALYSIS. DUE TO TIME CONSTRAINTS, AN UPDATED CODE WILL BE UPLOADED IN THE FUTURE.


#Repeat EWAS using lm() and for()


```{r error=FALSE, message=FALSE, warning=FALSE}

dat <- data.frame(dat)
df = data.frame(estimate = rep(NA,50), se = NA, tstatistic = NA, pvalue = NA)

test = lm(dat[,4]~dat$BMI)

summary(test)


```

```{r error=FALSE, message=FALSE, warning=FALSE. eval = FALSE}

as.numeric(coef(summary(test))[2,])

j = 1
df[j,] <- as.numeric(coef(summary(test))[2,])

```

```{r error=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
df = select(df, -starts_with("cpg_site"))
            
```

```{r error=FALSE, warning=FALSE, message=FALSE}
df = data.frame(estimate = rep(NA,50), se = NA, tstatistic = NA, pvalue = NA)

for(i in 3:54){
  test = lm(dat[,i]~dat$BMI)
  df[i,] = as.numeric(coef(summary(test))[2,])
}

# Placing the cpg names beside the results

df$cpg_site <- NA

df$cpg_site[3:54] = names(dat)[3:54]
df = na.omit(df)
```






# Modelling BMI over time


Linear mixed models

```{r error=FALSE, message=FALSE, warning=FALSE}

# Read in bmi data

bmi_lme_data <- fread("bmi_data.csv")

library(ggplot2)

# Plot bmi data over time
ggplot(bmi_lme_data, aes(x=age,y=bmi)) + geom_point(aes(color=as.factor(id))) +
  geom_smooth(method="lm", se = FALSE) +
  theme_bw() + theme(legend.position = "none")


ggplot(bmi_lme_data, aes(x=age,y=bmi)) +
  geom_smooth(se = FALSE) +
  theme_bw() + theme(legend.position = "none")

ggplot(bmi_lme_data, aes(x=age,y=bmi)) + geom_line(aes(color=as.factor(id))) +
  geom_smooth(method="lm", se = FALSE) +
  theme_bw() + theme(legend.position = "none")

library(tidyverse)
bmi_lme_data %>%
  filter(id<2000) %>%
ggplot(aes(x=age,y=bmi)) + geom_line(aes(color=as.factor(id))) +
  geom_smooth(method="lm", se = FALSE) +
  theme_bw() + theme(legend.position = "none")
```


```{r error=FALSE, message=FALSE, warning=FALSE}

# Model bmi data over time

library(nlme)

random_intercept_model <- lme(bmi~age, data = bmi_lme_data, random = ~1|id)

summary(random_intercept_model)

bmi_lme_data <- bmi_lme_data %>%
  mutate(age_spline7 = ifelse(age > 7, age - 7, 0),
         age_spline20 = ifelse(age > 20, age - 20, 0))

```

```{r error=FALSE, warning=FALSE, message=FALSE}

# Random Slope Model

random_slope_model <- lme(bmi ~ age, data = bmi_lme_data, random = ~age|id)

summary(random_slope_model)


# Random Slope and Spline Model
random_spline_model <- lme(bmi ~ age + age_spline7 + age_spline20, data = bmi_lme_data, random = ~age|id)

summary(random_spline_model)

anova(random_intercept_model, random_slope_model, random_spline_model)

```


EWAS aries BMI age 7

```{r}
bmi7 <- bmi_lme_data %>% filter(time == "F7")

```


```{r}

save(bmi_data_linked, resultF7, resultF9, resultF11, resultF12 , resultF13, resultF14, resultF15, resultF18, resultF24, file = "result.RData")

```


```{r}

save(cpg_assocF7,cpg_assocF9, cpg_assocF11, cpg_assocF12, cpg_assocF13, cpg_assocF14, cpg_assocF15, cpg_assocF18, cpg_assocF24, file = "cpg_assoc_results(F7-24).RData")

```
